---
title: "Maternal Mortality Ratio and Female Literacy"
output:
  word_document: default
  pdf_document: default
date: "2024-12-12"
---

```{r setup, include=FALSE}
library("readr")
library("dplyr")
library("data.table")
library("ggplot2")
library("writexl")
library("readxl")
```

## Maternal mortality and female literacy: an exploration and illustration

Maternal mortality ratios are an indicator and a function of both social and economic development. Literacy rates are an additional metric to measure socioeconomic progress and capital distributed across sex, location, and age. It is a well investigated interplay, that between MMR and female literacy rates [1]. Both measures might inform the other; therefore the role of social factors such as literacy in the development of broad-based social programs targeted towards lowering morbidity levels warrants a refocus in maternal mortality research. Maternal mortality explanations must take into account the socioeconomic factors such as income and literacy. 

Therefore, in this project we endeavor to wrangle the related data describing maternal mortality ratios and female literacy rates over time and location.

1. Merging data sets to build our own data frame for further analysis. By merging data sets from two different global health institutes we 1) bridge data structures and 2) highlight the data wrangling skills required for  and illustrate the dyanmics between MMR and female literacy rates in women ages 15+. 
2. Visualize and analyse the 


All of the code is located in the following GitHub repository: specific code chunks that derived the visualizations will be linked. 
https://github.com/isabellacolindres/HMS-520-Final-Project  


## Merging datasets

Code: https://github.com/isabellacolindres/HMS-520-Final-Project/pull/1 

The above code 1) uploads and 2) reconciles primary differences in data structure and nomenclature from datasets from two different sources.

1. Maternal mortality: https://ghdx.healthdata.org/record/ihme-data/maternal-mortality-estimates-and-mdg-5-attainment-country-1990-2011
2. Literacy rates for women: https://ourworldindata.org/grapher/adult-literacy-female.
rename columns and write functions to rename "global" to facilitate merging. Once that's complete, the datasets are merged by country and year and I write it into an excel output for next steps with visualizations and analysis.

We now have our data merged over time, location, and literacy rates (LR) and maternal mortality ratios (MMR). 

## Data structure and prep
```{r}
head(data)
```
The available rows and columns of overlapping data will allow us to examine MMR and female literacy at a global level, and to zoom into countries with 

## Global Landscape of MMR and Female Literacy
```{r}
setnames(data, "Literacy rate, adult female (% of females ages 15 and above)", "LR") #easier nomenclature for plotting

global_scale <- subset(data, Country == "GLOBAL") #subsetting to global

ggplot(global_scale, aes(x = Year, y = `Maternal mortality ratio (MMR)`)) +
  geom_point() +
  theme_minimal() +
  geom_smooth(method = "lm", se = TRUE, color = "purple") +
  labs(title = "Global MMR over Years 1990-2011")
```
```{r}
ggplot(global_scale, aes(x = Year, y = LR)) +
  geom_point() +
  theme_minimal() +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(title = "Global Female Literacy Rate (%) over Years 1990-2011",
       y = "Female literacy rate % ages above 15")
```

The global MMR demonstrates a steady decline between the years 1990 and 2011, in parallel to the female literacy rates which demonstrate a steady global increase amongst women ages 15 and above. 








## Sources
[1] Pillai, Vijayan K., Maleku, Arati, Wei, FangHsun, Maternal Mortality and Female Literacy Rates in Developing Countries during 1970–2000: A Latent Growth Curve Analysis, International Journal of Population Research, 2013, 163292, 11 pages, 2013. https://doi.org/10.1155/2013/163292
